Categories of Optics

### 2018-05-19 paper.tex
Commit things before rekerjiggering

%\todo{Past work:
%\begin{itemize}
%\item The Haskell lens people: edwardk, shachaf, xplat, roconnor... (TODO: find out who these people are)
%\item The Doubles paper
%\item Bartosz's blog post which was written about the same time I started this. (Only tensor-like optics, no discussion of laws, confusion about enrichment(?))
%\item The profunctor optics paper out of Oxford. (No generality, nothing categorical, no discussion of laws)
%\end{itemize}
%}
%
%\todo{Contributions:
%\begin{itemize}
%\item ...
%\end{itemize}
%}
%
%\todo{Introduction should include:
%\begin{itemize}
%\item Previous attempts to generalise optics
%\item Some remark about how it's hard to isolate exactly what $\Set$-like properties are being used
%\end{itemize}
%}
A lens from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$ that typically obeys three laws (whose names are in diagrammatic order):
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be \todo{represented faithfully in $S$},
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times \id_A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}

@@

@@

In favourable conditions, we can upgrade this result to show that the existence of a lens $S \hto A$ implies that $S \cong M \times A$ for some $M$. This is a generalised from the argument given in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

@@

@@

\begin{align*}
\Optic_\otimes((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \\
&\cong \int^{M \in \C} \C(S, M \otimes A) \times \C(M, \homC(A',S')) \\
&\cong \C(S, \homC(A',S') \otimes A)
\end{align*}
where $\homC(S, A)$ denotes the internal hom. If $\C$ is cartesian, this is of course isomorphic to the pair of maps discussed in the previous section.

@@

@@

\begin{corollary}
A profunctor optic is determined by its component at $\Pastro_\M E_A$, and furthermore, this component is determined by its value on $\langle \id_A, \id_A, \id_A, \id_A \rangle \in (U \Pastro_\M E_A)(A, A)$.
\end{corollary}
\begin{proof}
This is the content of the first two isomorphisms above.

Explicitly, let $p_{\Pastro_\M E_A}(\langle \id_A, \id_A, \id_A, \id_A \rangle) = \langle l, \id_A, \id_A, r \rangle$, where $l : S \to M A$ and $r : M A \to S$. Then the component of $p$ at any other Tambara module $P$ is given by
\begin{align*}
p_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
where $\alpha$ is the module structure for $P$.
\end{proof}

@@

\begin{proposition}
If $p : S \hto A$ is a lawful optic then the associated natural transformation $\tilde{p} : (U-)(A,A) \to (U-)(S,S)$ is a lax monoidal.
\end{proposition}
\begin{proof}
Let $\langle l, r \rangle$ be a representative for $p$ in $\Optic_\M(S,A)$ via the isomorphism of Proposition \ref{prop-profunctor-optics-are-optics}, so that
\begin{align*}
\tilde{p}_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
on each Tambara module $P$.

>>>>

%\todo{Past work:
%\begin{itemize}
%\item The Haskell lens people: edwardk, shachaf, xplat, roconnor... (TODO: find out who these people are)
%\item The Doubles paper
%\item Bartosz's blog post which was written about the same time I started this. (Only tensor-like optics, no discussion of laws, confusion about enrichment(?))
%\item The profunctor optics paper out of Oxford. (No generality, nothing categorical, no discussion of laws)
%\end{itemize}
%}
%
%\todo{Contributions:
%\begin{itemize}
%\item ...
%\end{itemize}
%}
%
%\todo{Introduction should include:
%\begin{itemize}
%\item Previous attempts to generalise optics
%\item Some remark about how it's hard to isolate exactly what $\Set$-like properties are being used
%\end{itemize}
%}
A lens from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$ that typically obeys three laws (whose names are in diagrammatic order):
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be represented faithfully in $S$;
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times \id_A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}

@@

The general pattern is, once we choose a particular monoidal action $\M \to [\C, \C]$, we can find an isomorphism between the set of optics $(S, S') \hto (A, A')$ and a set that is easier to describe. The goal is then to find necessary and sufficient conditions on this second set for the corresponding element of the coend to be a lawful optic.

@@

In favourable conditions, we can upgrade this result to show that the existence of a lens $S \hto A$ implies that $S \cong C \times A$ for some $C$. This is a generalised from the argument given in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

@@

\todo{There may be something generalisable here with ``monoidal factorisation systems''}

@@

\begin{align*}
\Optic_\otimes((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \\
&\cong \int^{M \in \C} \C(S, M \otimes A) \times \C(M, \homC(A',S')) \\
&\cong \C(S, \homC(A',S') \otimes A)
\end{align*}
where $\homC(A', S')$ denotes the internal hom. If $\C$ is cartesian, this is of course isomorphic to the pair of maps discussed in the previous section.

@@

Suppose $\C$ is cartesian closed and has finite coproducts. Then $\C$ acts on itself via $M \cdot A = (M \sqcup 1) \times A$.
\begin{align*}
\Optic((S, S'), (A, A')) 
&= \int^{M \in \C} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
&= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
&= \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
&= \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S') \\
\end{align*}

\todo{Equivalently(?), let $\M$ be the category of pointed sets}

@@

For $p : S \hto A$, let $\tilde{p} : (U-)(A,A) \Rightarrow (U-)(S,S)$ denote the corresponding natural transformation under this isomorphism, and for $t : (U-)(A,A) \Rightarrow (U-)(S,S)$, let $\hat{t} : S \hto A$ be the corresponding optic.

\begin{corollary}
A profunctor optic is determined by its component at $\Pastro_\M E_A$, and furthermore, this component is determined by its value on $\langle \id_A, \id_A, \id_A, \id_A \rangle \in (U \Pastro_\M E_A)(A, A)$.
\end{corollary}
\begin{proof}
This is the content of the first two isomorphisms above.
\todo{fix notation in the following}
Explicitly, let $p_{\Pastro_\M E_A}(\langle \id_A, \id_A, \id_A, \id_A \rangle) = \langle l, \id_A, \id_A, r \rangle$, where $l : S \to M A$ and $r : M A \to S$. Then the component of $p$ at any other Tambara module $P$ is given by
\begin{align*}
p_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
where $\alpha$ is the module structure for $P$.
\end{proof}

@@

\begin{proposition}
If $p : S \hto A$ is a lawful optic then the associated natural transformation $\tilde{p} : (U-)(A,A) \to (U-)(S,S)$ is lax monoidal.
\end{proposition}
\begin{proof}
Let $\langle l, r \rangle$ be a representative for $p$ in $\Optic_\M(S,A)$ via the isomorphism of Proposition \ref{prop-profunctor-optics-are-optics}, so that
\begin{align*}
\tilde{p}_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
on each Tambara module $P$.
